\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-12 paper/ms.tex
more edits to appendix

\subsection{Scalable computation of celerite models}

The \celeriteterm\ model described in \sect{celerite} can be exactly
represented as a rank $R = 2\,J$ semiseparable matrix\footnote{We note that if
some \celeriteterm\ terms are real ($b_j=0$ and $d_j=0$) then these terms only
add one to the rank.
Therefore, if a \celeriteterm\ model has $J_\mathrm{real}$ real and
$J_\mathrm{complex}$ general terms, the semiseparable rank is only
$R=J_\mathrm{real} + 2\,J_\mathrm{complex}=2\,J-J_{real}$.}
where the components of the relevant matrices are
\begin{eqnarray}\eqlabel{system}
U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \nonumber\\
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \nonumber\\
V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \nonumber\\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \nonumber\\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}
A na\"ive implementation of the algorithm in \eq{algorithm} for this system
will result in numerical overflow and underflow issues in many realistic cases
because both $U$ and $V$ have factors of the form $e^{\pm c_j t}$, where $t$
can be any arbitrarily large real number.
This issue can be avoided by reparameterizing the semiseparable representation
from \eq{system}.
%A similar issue was also noted by \citet{Ambikasaran:2015} and it can be
%avoided using a reparameterization.
We define the following pre-conditioned variables
\begin{eqnarray}\eqlabel{reparam}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \nonumber\\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \nonumber\\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \nonumber\\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \nonumber\\
\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \nonumber\\
\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
\end{eqnarray}
and the new set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm to compute the Cholesky factorization of a \celeriteterm\ model
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}
As before, a single iteration of this algorithm requires $\mathcal{O}(J^2)$
operations and it is repeated $N$ times for a total scaling of
$\mathcal{O}(N\,J^2)$.
Furthermore, $\tilde{V}_{n,j}$ and $A_{n,n}$ can be updated in place and used
to represent $\tilde{W}_{n,j}$ and $D_{n,n}$, giving a total memory footprint
of $(6\,J + 1)\,N + J\,(J-1)/2$ floating point numbers.

@@

\newpage
\appendix

@@

\subsection{Multiplication and filtering}

The product of a rank-$R$ semiseparable matrix with general vectors and
matrices can be computed in $\mathcal{O}(N\,R)$ operations
\citep{Vandebril:2007}.
In the context of \celeriteterm\ models, na\"ive application of these methods
will result in numerical overflow and underflow issues much like we found with
the Cholesky factorization.
To avoid these errors, we derive the following numerically stable algorithm
for computing the dot product of a \celeriteterm\ covariance matrix
\begin{eqnarray}
\bvec{y} = K\,\bvec{z}
\end{eqnarray}
using the semiseparable representation for $K$ defined in \eqalt{reparam}:
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j}\,\left(f^+_{n+1,j} + \tilde{U}_{n,j}\,z_{n+1}\right) \\
f^-_{n,j} &=& \phi_{n,j}\,\left(f^-_{n-1,j} + \tilde{V}_{n-1,j}\,z_{n-1}\right) \\
y_n &=& A_{n,n}\,z_n + \sum_{j=1}^{2\,J} \left[ \tilde{V}_{n,j} f^+_{n,j} +
    \tilde U_{n-1,j} f^-_{n,j}\right] \quad,
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$.
This algorithm requires two sweeps~--~$n=2,\,\ldots,\,N$ to compute $f^-$ and
$n=N-1,\,\ldots,\,1$ to compute $f^+$~--~with a scaling of $\mathcal{O}(N\,J)$.

\subsection{Simulation}

Simulation of a Gaussian process for large $N$ requires multiplying an
$N$-dimensional vector of Gaussian normal deviates, $\bvec{q}$, by the square root
of the covariance matrix:
\begin{equation}
\bvec{y} = K_\bvec{\alpha}^{1/2} \bvec{q},
\end{equation}
where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the first term of $\mathcal{Q} = -2 \mathcal{L}$
with $\mu_\bvec{\theta} =0$ is given by
\begin{eqnarray}
\mathcal{Q} &=&\bvec{y}^T K_\bvec{\alpha}^{-1} \bvec{y}\\
&=& \bvec{y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} \bvec{y}\\
&=& \bvec{q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} \bvec{q}\\
&=& \bvec{q}^T \bvec{q}.
\end{eqnarray}
The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$; for
diagonal $K_\bvec{\alpha}$, $\mathcal{Q}$ becomes the $\chi^2$ statistic.

To generate a draw from a Gaussian process with a specified \celerite\ model requires carrying out the
low-rank Cholesky decomposition first, generating the vector $\bvec{q}$ of random normal deviates, and then multiplying
\begin{equation}
\bvec{y} = L D^{1/2} \bvec{q}.
\end{equation}
This may be carried out in $\mathcal{O}(JN)$ operations with
\begin{eqnarray}
f_{n,j} &=& \phi_{n,j} \left(f_{n-1,j} + \tilde{W}_{n-1,j} D_{n-1,n-1}^{1/2} q_{n-1}\right)\\
y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2\,J} \tilde{U}_{n,k} f_{n,k}
\end{eqnarray}
for $n = 1$ to $N$ and $j=1$ to $2\,J$, with $f_{0,j} = 0$ for all $j$ and $q_0 = 0$.
We expect that this simulation algorithm will be useful for test-driving Gaussian
process inference algorithms applied to large datasets as even simulation of
Gaussian processes becomes prohibitive for large $N$ if standard Cholesky decomposition
and multiplication are used.

\subsection{Prediction and interpolation}

%Gaussian process regression may also be used for prediction and interpolation at future or missing data, based upon current measurements.  With \celerite\ it proves possible to predict the mean value at times ${t_m; m=1,...,M}$ based upon measured times ${t_n; n=1,...,N}$ in $\mathcal{O}(N+M)$ operations.

For predicting or interpolating missing or future data points (and their variance) at times $\bvec{t^*} = \left\{t_{1}^*,t_{2}^*,...,t_{M}^*\right\}$, we need to apply the covariance between the measured and predicted times.  The mean of the predicted values is (equation 2.19 from Rasmussen \& Williams):
\begin{eqnarray}
\bar f^* &=& K(\bvec{t^*},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y},\\
&=& K(\bvec{t^*},\bvec{t}) \bvec{z}
\end{eqnarray}
where $\bvec{y}$ is the vector of measurements and $\sigma$ is the measurement uncertainty, and
$\bvec{z} = \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y}$ can be computed from
the Cholesky decomposition of the full covariance matrix in ${\cal O}(J^2N)$.

Since the covariance matrix may be written as the sum of exponentials, the predicted mean is given by
\begin{eqnarray}
\bar f^*_m &=& K(t^*_m,\bvec{t}) \bvec{z} \\
&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t^*_m-t_n\vert}\left(a_j \cos{(d_j \vert t^*_m-t_n\vert)} +b_j \sin{(d_j \vert t^*_m-t_n\vert)}\right) z_j,
\end{eqnarray}
for $m =1$ to $M$.
Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t^*_m$ and
$\{t_{n0+1},...,t_N\} \ge t^*_m$, gives
\begin{eqnarray}
\bar f^*_m &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t^*_m-t_n)} \left[a_j \cos{(d_j(t^*_m-t_n))}+b_j\sin{(d_j(t^*_m-t_n))}\right] z_n \\
&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t^*_m)} \left[a_j \cos{(d_j(t_n-t^*_m))}+b_j\sin{(d_j(t_n-t^*_m))}\right] z_n.
\end{eqnarray}
We can compute this recursively.  Defining
\begin{eqnarray}
Q^-_{n,i} &=& \left(Q^-_{n-1,i}+z_n\tilde V_{n,i}\right)e^{-c_j(t_{n+1}-t_{n})},\\
X^-_{m,n,i} &=& e^{-c_j(t_m^*-t_{n+1})}\tilde U_{m,i}^*,\\
Q^+_{n,i} &=& \left(Q^+_{n+1,i}+z_n\tilde U_{n,i}\right)e^{-c_j(t_{n}-t_{n-1})},\\
X^+_{m,n,i} &=& e^{-c_j(t_{n-1}-t_m^*)} \tilde V_{m,i}^*,
\end{eqnarray}
where $t_0=t_1$, $t_{N+1}=t_N$, $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$
to $2\,J$, and in $X^\pm_{m,n,i}$ the expressions
for $\tilde U_{m,i}^*$ and $\tilde V_{m,i}^*$ are evaluted at $t_m^*$.
Two sweeps must be carried out, upward in $n$ and $m$ to compute $Q^-$ recursively, and downward in
$m$ and $n$ to compute $Q^+$ recursively, for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.
For each value of $m$, $Q^\pm$ are recursively updated over $n$ until $n_0$ is reached, at which
point the predicted mean can be computed from
\begin{equation}
\bar f^*_i = \sum_{i=1}^{2\,J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0+1,i}X^+_{m,n_0+1,i}\right],
\end{equation}

%The variance evaluated at future predicted times $\bvec{t^*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
%\begin{eqnarray}
%\mathrm{var}(f^*_{i+1}) &=& K(t^*_{i+1},t^*_{i+1}) - K(t^*_{i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t^*_{i+1})\\
%&=& K(t^*_{i+1},t^*_{i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t^*_{i+1}-t^*_{i})} \\
%&\times& e^{-\beta_p(t^*_{i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t^*_{i}-\bvec{t})}\\
%&=& K(t^*_{i+1},t^*_{i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t^*_{i+1}-t^*_{i})} P_{i,p,q}\\
%\end{eqnarray}
%for $1 \le p,q \le J$,
%where
%\begin{equation}
%P_{i,p,q} = e^{-\beta_p(t^*_{i}-\bvec{t}^T)}
%K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t^*_{i}-\bvec{t})}.
%\end{equation}
%For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
%with the Cholesky decomposition.
%Subsequent $P_{i+1,p,q}$ may be obtained recursively as
%\begin{equation}
%P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t^*_{i+1}- t^*_{i})} P_{i,p,q},
%\end{equation}
%in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
%Note that this variance formula {\it only} works for prediction; we have
%yet to derive a formula for which $\bvec{t^*}$ and $\bvec{t}$ are
%interspersed.
}

\newpage
\bibliography{celerite}

>>>>

\subsection{Scalable computation of celerite models}\sectlabel{scalable}

The \celeriteterm\ model described in \sect{celerite} can be exactly
represented as a rank $R = 2\,J$ semiseparable matrix\footnote{We note that if
some \celeriteterm\ terms are real ($b_j=0$ and $d_j=0$) then these terms only
add one to the rank.
Therefore, if a \celeriteterm\ model has $J_\mathrm{real}$ real and
$J_\mathrm{complex}$ general terms, the semiseparable rank is only
$R=J_\mathrm{real} + 2\,J_\mathrm{complex}=2\,J-J_{real}$.}
where the components of the relevant matrices are
\begin{eqnarray}\eqlabel{system}
U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \nonumber\\
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \nonumber\\
V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \nonumber\\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \nonumber\\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}
A na\"ive implementation of the algorithm in \eq{algorithm} for this system
will result in numerical overflow and underflow issues in many realistic cases
because both $U$ and $V$ have factors of the form $e^{\pm c_j t}$, where $t$
can be any arbitrarily large real number.
This issue can be avoided by reparameterizing the semiseparable representation
from \eq{system}.
%A similar issue was also noted by \citet{Ambikasaran:2015} and it can be
%avoided using a reparameterization.
We define the following pre-conditioned variables
\begin{eqnarray}\eqlabel{reparam}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \nonumber\\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \nonumber\\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \nonumber\\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \nonumber\\
\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \nonumber\\
\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
\end{eqnarray}
and the new set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm to compute the Cholesky factorization of a \celeriteterm\ model
\begin{eqnarray}\eqlabel{cholesky-factor}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}
As before, a single iteration of this algorithm requires $\mathcal{O}(J^2)$
operations and it is repeated $N$ times for a total scaling of
$\mathcal{O}(N\,J^2)$.
Furthermore, $\tilde{V}_{n,j}$ and $A_{n,n}$ can be updated in place and used
to represent $\tilde{W}_{n,j}$ and $D_{n,n}$, giving a total memory footprint
of $(6\,J + 1)\,N + J\,(J-1)/2$ floating point numbers.

@@

\vspace{5ex}
\appendix

@@

\subsection{Multiplication}

The product of a rank-$R$ semiseparable matrix with general vectors and
matrices can be computed in $\mathcal{O}(N\,R)$ operations
\citep{Vandebril:2007}.
In the context of \celeriteterm\ models, na\"ive application of these methods
will result in numerical overflow and underflow issues much like we found with
the Cholesky factorization.
To avoid these errors, we derive the following numerically stable algorithm
for computing the dot product of a \celeriteterm\ covariance matrix
\begin{eqnarray}
\bvec{y} = K\,\bvec{z}
\end{eqnarray}
using the semiseparable representation for $K$ defined in \eqalt{reparam}:
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j}\,\left[f^+_{n+1,j} +
    \tilde{U}_{n,j}\,z_{n+1}\right] \\
f^-_{n,j} &=& \phi_{n,j}\,\left[f^-_{n-1,j} +
    \tilde{V}_{n-1,j}\,z_{n-1}\right] \\
y_n &=& A_{n,n}\,z_n + \sum_{j=1}^{2\,J} \left[ \tilde{V}_{n,j} f^+_{n,j} +
    \tilde U_{n-1,j} f^-_{n,j}\right] \quad,
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$.
This algorithm requires two sweeps~--~$n=2,\,\ldots,\,N$ to compute $f^-$ and
$n=N-1,\,\ldots,\,1$ to compute $f^+$~--~with a scaling of $\mathcal{O}(N\,J)$.

\subsection{Sampling data from a celerite process}

To sample a dataset $\bvec{y}$ from a GP model for fixed parameters
$\bvec{\theta}$ and $\bvec{\alpha}$, we must compute
\begin{equation} \eqlabel{sim}
\bvec{y} = \bvec{\mu}_\bvec{\theta}(\bvec{t})+
    K_\bvec{\alpha}^{1/2}\,\bvec{q}
\end{equation}
where $\bvec{\mu}_\bvec{\theta}(\bvec{t})$ is the mean function evaluated at
the input coordinates $\bvec{t}$, $\bvec{q}$ is a vector of draws from the
unit normal $q_i \sim \mathcal{N}(0,\,1)$, and
\begin{eqnarray}
K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \,
    {\left(K_\bvec{\alpha}^{1/2}\right)}^T\quad.
\end{eqnarray}
For a general matrix $K_\bvec{\alpha}$, the cost of this operation scales as
$\mathcal{O}(N^3)$ to compute $K_\bvec{\alpha}^{1/2}$ and $\mathcal{O}(N^2)$
to compute the dot product.
For \celeriteterm\ models, we use the semiseparable representation of the
Cholesky factor that we derived in \sect{cholesky} to compute the dot product
\eq{sim} in $\mathcal{O}(N\,J)$.
Using the notation from \sect{scalable}, the algorithm to compute this dot
product is
\begin{eqnarray}
f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} + \tilde{W}_{n-1,j}\,
    \sqrt{D_{n-1,n-1}}\,q_{n-1}\right]\\
y_n &=& \sqrt{D_{n,n}}\,q_n + \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
\end{eqnarray}
where $f_{1,j} = 0$ for all $j$.
This scalable algorithm is useful for generating large simulated
datasets for experiments with \celerite\ and for performing posterior
predictive checks.

\subsection{Interpolation \& extrapolation}

The predictive distribution of a GP at a vector of $M$ input coordinates
\begin{eqnarray}
\bvec{y}^* = \left(\begin{array}{ccccc}
    t_1^*\quad && \cdots\quad && t_M^*
\end{array}\right)^\T
\end{eqnarray}
conditioned on a dataset $(\bvec{t},\,\bvec{y})$ and GP
parameters $(\bvec{\theta},\,\bvec{\alpha})$, is a normal $\bvec{y}^* \sim
\mathcal{N}(\bvec{\mu}^*,\,K^*)$ with mean
\begin{eqnarray}\eqlabel{pred-mean}
\bvec{\mu}^* &=& \bvec{\mu}_\bvec{\theta}(\bvec{t}^*) +
    K(\bvec{t}^*,\,\bvec{t})\,{K(\bvec{t},\,\bvec{t})}^{-1}\,\left[\bvec{y} -
    \bvec{\mu}_\bvec{\theta}(\bvec{t})\right]
\end{eqnarray}
and covariance
\begin{eqnarray}
K^* &=& K(\bvec{t}^*,\,\bvec{t}^*) -
    K(\bvec{t}^*,\,\bvec{t})\,{K(\bvec{t},\,\bvec{t})}^{-1}\,
    K(\bvec{t},\,\bvec{t}^*)
\end{eqnarray}
where $K(\bvec{v},\,\bvec{w})$ is the covariance matrix computed between
\bvec{v} and \bvec{w} \citep{Rasmussen:2006}.

Na\"ively, the computational cost of computing \eq{pred-mean} for a
\celeriteterm\ model scales as $\mathcal{O}(N\,M)$ if we reuse the Cholesky
factor derived in \sect{cholesky}.
It is, however, possible to improve this scaling to $\mathcal{O}(n\,N +
m\,M)$ where $n$ and $m$ are integer constants.
To derive this, we expand \eq{pred-mean} using \eq{celerite-kernel}
\begin{eqnarray}
\mu^*_m &=& K(t^*_m,\bvec{t}) \bvec{z} \\
    &=& \sum_{n=1}^N \sum_{j=1}^J e^{-c_j \vert t^*_m-t_n\vert}\,\left[
        a_j\,\cos{(d_j\,\vert t^*_m-t_n\vert)} +
        b_j\,\sin{(d_j\,\vert t^*_m-t_n\vert)}
    \right] \,z_{n} \quad. \nonumber
\end{eqnarray}
where
\begin{eqnarray}
\bvec{z} = {K(\bvec{t},\,\bvec{t})}^{-1}\,\left[\bvec{y} -
    \bvec{\mu}_\bvec{\theta}(\bvec{t})\right] \quad.
\end{eqnarray}
Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t^*_m$ and
$\{t_{n0+1},...,t_N\} \ge t^*_m$, gives
\begin{eqnarray}
\mu^*_m &=& \sum_{j=1}^J\sum_{n=1}^{n_0}
    e^{-c_j\,(t^*_m-t_n)}\,\left[a_j\,\cos{(d_j\,(t^*_m-t_n))}
        +b_j\,\sin{(d_j\,(t^*_m-t_n))}\right]\,z_n \\
&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j\,(t_n-t^*_m)}\,\left[
    a_j\,\cos{(d_j\,(t_n-t^*_m))}+b_j\,\sin{(d_j\,(t_n-t^*_m))}\right]\,z_n
\quad. \nonumber
\end{eqnarray}
We compute this using two passes, a forward pass $n_0 = 1,\,\ldots,\,N$ and
a backward pass $n_0 = N,\,\ldots,\,1$.
Defining
\begin{eqnarray}
Q^-_{n,k} &=& \left[Q^-_{n-1,k}+z_n\,\tilde V_{n,k}\right]\,e^{-c_{k//2}\,(t_{n+1}-t_{n})},\\
X^-_{m,n,k} &=& e^{-c_{k//2}\,(t_m^*-t_{n+1})}\,\tilde U_{m,k}^*,\\
Q^+_{n,k} &=& \left[Q^+_{n+1,k}+z_n\,\tilde U_{n,k}\right]\,e^{-c_{k//2}(t_{n}-t_{n-1})},\\
X^+_{m,n,k} &=& e^{-c_{k//2}\,(t_{n-1}-t_m^*)}\,\tilde V_{m,k}^* \quad,
\end{eqnarray}
where $t_0=t_1$, $t_{N+1}=t_N$, $Q^-_{0,k} = 0$ and $Q^+_{N+1,k}=0$ for $k=1$
to $2\,J$, and in $X^\pm_{m,n,k}$ the expressions
for $\tilde U_{m,i}^*$ and $\tilde V_{m,i}^*$ are evaluated at $t_m^*$.
For each value of $m$, $Q^\pm$ are recursively updated over $n$ until $n_0$ is
reached, at which point the predicted mean can be computed from
\begin{equation}
\mu^*_m = \sum_{k=1}^{2\,J}
    \left[Q^-_{n_0,k}X^-_{m,n_0,k}+Q^+_{n_0+1,k}X^+_{m,n_0+1,k}\right] \quad.
\end{equation}
}

\bibliography{celerite}
